---
title: "Endogenous cells dataset, preliminary data analysis, cluster resolution `r params$cluster.resolution`"
author:
  - "trnguyen@ukaachen.de"
date: "Last update on `r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    keep_md: no
    df_print: paged
    toc: true
    toc_float:
      toc_collapsed: false
    toc_depth: 3
    number_sections: true
    theme: lumen
params:
  cluster.resolution: NA
---


```{css zoom-lib-src, echo = FALSE}
script src = "https://ajax.googleapis.com/ajax/libs/jquery/3.4.1/jquery.min.js"
```

```{js zoom-jquery, echo = FALSE}
 $(document).ready(function() {
    $('body').prepend('<div class=\"zoomDiv\"><img src=\"\" class=\"zoomImg\"></div>');
    // onClick function for all plots (img's)
    $('img:not(.zoomImg)').click(function() {
      $('.zoomImg').attr('src', $(this).attr('src')).css({width: '100%'});
      $('.zoomDiv').css({opacity: '1', width: 'auto', border: '1px solid white', borderRadius: '5px', position: 'fixed', top: '50%', left: '50%', marginRight: '-50%', transform: 'translate(-50%, -50%)', boxShadow: '0px 0px 50px #888888', zIndex: '50', overflow: 'auto', maxHeight: '100%'});
    });
    // onClick function for zoomImg
    $('img.zoomImg').click(function() {
      $('.zoomDiv').css({opacity: '0', width: '0%'}); 
    });
  });
```

<style type="text/css">
    div.datatables { height: auto !important;}
</style>


```{r echo=FALSE, warning=FALSE, message=FALSE, results='hide', fig.width=20, fig.height=12}
#####----------------------------------------------------------------------#####
#
# trnguyen@ukaachen.de
#
#####----------------------------------------------------------------------#####

##### clean up #####
# gc()
# rm(list = ls())

scrna_pipeline_src <- "/home/hieunguyen/CRC1382/src_2023/src_pipeline/scRNA_GEX_pipeline/processes_src"
source(file.path(scrna_pipeline_src, "import_libraries.R"))
source(file.path(scrna_pipeline_src, "helper_functions.R"))

source(file.path(scrna_pipeline_src, "s8_integration_and_clustering.R"))

#####----------------------------------------------------------------------#####
# CONFIGURATIONS 
#####----------------------------------------------------------------------#####
chosen.seed <- 42
num.dim.integration <- 25 
num.PCA <- 25
num.dim.cluster <- 25
num.PC.used.in.Clustering <- 25
num.PC.used.in.UMAP <- 25
my_random_seed <- 42

path.to.src <- "/home/hieunguyen/CRC1382/src_2023/src_pipeline/scRNA_GEX_pipeline/processes_src"

outdir <- "/home/hieunguyen/CRC1382/outdir"
PROJECT <- "SAlBounny_full"
output.version <- "20241021"

if ("svglite" %in% installed.packages() == FALSE){
  install.packages("svglite")
}
path.to.main.output <- file.path(outdir, PROJECT, output.version, "data_analysis")
dir.create(path.to.main.output, showWarnings = FALSE, recursive = TRUE)

path.to.01.output <- file.path(path.to.main.output, "01_output")
path.to.02.output <- file.path(path.to.main.output, "02_output")
path.to.03.output <- file.path(path.to.main.output, "03_output")
dir.create(path.to.03.output, showWarnings = FALSE, recursive = TRUE)

s.obj.origin <- readRDS(file.path(path.to.02.output, "SAlBounny_full.filterCD4.rds"))

##### remove some clusters of contaminated cells
s.obj <- subset(s.obj.origin, seurat_clusters %in% c(7, 9, 10, 12, 13, 14) == FALSE)

# cluster.resolution <- params$cluster.resolution
cluster.resolution <- 0.5
```


```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
if (file.exists(file.path(path.to.03.output, sprintf("%s.filter_contaminated_cells.clusterRes_%s.rds", PROJECT, cluster.resolution))) == FALSE){
  num.PCA <- 30
  num.PC.used.in.UMAP <- 30
  num.PC.used.in.Clustering <- 30
  num.dim.integration <- 30
  num.dim.cluster <- 30
  
  DefaultAssay(s.obj) <- "RNA"
  s.obj <- s8.integration.and.clustering(s.obj, 
                             path.to.output = NULL, 
                             save.RDS.s8 =  FALSE,
                             PROJECT = NULL, 
                             num.dim.integration = num.dim.integration ,
                             num.PCA = num.PCA,
                             num.PC.used.in.UMAP = num.PC.used.in.UMAP,
                             num.PC.used.in.Clustering = num.PC.used.in.Clustering,
                             cluster.resolution = cluster.resolution,
                             my_random_seed = 42,
                             umap.method = "uwot",
                             genes.to.not.run.PCA = NULL,
                             inte_pca_reduction_name = "INTE_PCA", 
                             inte_umap_reduction_name = "INTE_UMAP")
  saveRDS(object = s.obj, file.path(path.to.03.output, sprintf("%s.filter_contaminated_cells.clusterRes_%s.rds", PROJECT, cluster.resolution)))
} else {
  s.obj <- readRDS(file.path(path.to.03.output, sprintf("%s.filter_contaminated_cells.clusterRes_%s.rds", PROJECT, cluster.resolution)))
}
```

# UMAP all cells
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
DimPlot(object = s.obj, reduction = "INTE_UMAP", label = TRUE, label.box = TRUE, repel = TRUE)
```


# Cluster marker genes

Applying the `wilcox` test on all genes to find the highly expressed genes in each cluster. 

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
if (file.exists(file.path(path.to.03.output, sprintf("cluster_marker_genes_clusterRes_%s.rds", cluster.resolution))) == FALSE){
  cluster.markers <- FindAllMarkers(object = s.obj, assay = "RNA", test.use = "wilcox")
  cluster.markers <- subset(cluster.markers, cluster.markers$p_val_adj <= 0.05 & cluster.markers$avg_log2FC >= 0)
  saveRDS(cluster.markers, file.path(path.to.03.output, sprintf("cluster_marker_genes_clusterRes_%s.rds", cluster.resolution))  )
} else {
  cluster.markers <- readRDS(file.path(path.to.03.output, sprintf("cluster_marker_genes_clusterRes_%s.rds", cluster.resolution)))
}

```


### Heatmap: Top N up-gene in each clusters
```{r echo=FALSE, warning=FALSE, results='asis', message=FALSE, fig.width=15, fig.height=10}
topN <- 10
tmp.cluster.markers.up <- subset(cluster.markers, cluster.markers$avg_log2FC >= 0 & cluster.markers$p_val_adj <= 0.05)

selected.genes <- data.frame()
for (cluster.id in unique(tmp.cluster.markers.up$cluster)){
  tmp <- subset(tmp.cluster.markers.up, tmp.cluster.markers.up$cluster == cluster.id) %>% arrange(desc(avg_log2FC))
  tmp.genes <- head(tmp, topN)
  selected.genes <- rbind(selected.genes, tmp.genes)
}

p <- DoHeatmap(object = s.obj, features = selected.genes$gene) + scale_fill_viridis_c(option = "plasma")
ggsave(plot = p, filename = "Heatmap_topN_up_genes_in_all_clusters.svg", path = file.path(path.to.03.output), device = "svg", width = 28, height = 20, dpi = 300)
```


```{r echo=FALSE, warning=FALSE, results='asis', message=FALSE, fig.width=28, fig.height=20}
print(p)
```


### Tables of all cluster marker genes {.tabset}
```{r echo=FALSE, include=FALSE}
cluster.markers %>% create_dt()
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=15}

for (cluster.id in sort(unique(cluster.markers$cluster))){
  tmp.cluster.markers <- subset(cluster.markers, cluster.markers$cluster == cluster.id) %>% arrange(desc(avg_log2FC))
  tmp.cluster.markers <- tmp.cluster.markers %>% mutate_if(is.numeric, round, 6) 
  
  cat(paste("\n\n#### Cluster: ", cluster.id, "##\n"))
  print( htmltools::tagList(datatable(tmp.cluster.markers, extensions = 'Buttons',
                                      filter = "top",
                                      options = list(dom = 'Blfrtip',
                                                     buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
                                                     lengthMenu = list(c(10,25,50,-1),
                                                                       c(10,25,50,"All")),
                                                     columnDefs = list(list(
                                                       targets = "_all",
                                                       render = JS(
                                                         "function(data, type, row, meta) {",
                                                         "return type === 'display' && data != null && data.length > 100 ?",
                                                         "'<span title=\"' + data + '\">' + data.substr(0, 100) + '...</span>' : data;",
                                                         "}")
                                                     ))
                                      ))))
  cat("\n \n")
}

```


### Top-12 cluster marker genes `FeaturePlot` {.tabset}
Here we just show the top-12 genes (ranked by `avg_log2FC`) for each cluster. 

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=15}
for (cluster.id in sort(unique(s.obj@meta.data$seurat_clusters))){
  DefaultAssay(s.obj) <- "RNA"
  cat(sprintf("#### Cluster %s \n", cluster.id))
  tmp.cluster.markers <- subset(cluster.markers, cluster.markers$cluster == cluster.id) %>% arrange(desc(avg_log2FC))
  p <- FeaturePlot(object = s.obj, reduction = "INTE_UMAP", features = head(tmp.cluster.markers, 12)$gene, ncol = 3, label = TRUE, pt.size = 0.5, label.size = 5, label.color = "red")  
  print(p)
  cat("\n \n")
}
```

### Top-12 cluster marker genes `DotPlot` {.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=15}
for (cluster.id in sort(unique(s.obj@meta.data$seurat_clusters))){
  DefaultAssay(s.obj) <- "RNA"
  cat(sprintf("#### Cluster %s \n", cluster.id))
  tmp.cluster.markers <- subset(cluster.markers, cluster.markers$cluster == cluster.id) %>% arrange(desc(avg_log2FC))
  p <- DotPlot(object = s.obj, features = head(tmp.cluster.markers, 9)$gene)  
  print(p)
  cat("\n \n")
}
```

### Top-12 cluster marker genes `Vlnplot` {.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=30, fig.height=20}
for (cluster.id in sort(unique(s.obj@meta.data$seurat_clusters))){
  DefaultAssay(s.obj) <- "RNA"
  cat(sprintf("#### Cluster %s \n", cluster.id))
  tmp.cluster.markers <- subset(cluster.markers, cluster.markers$cluster == cluster.id) %>% arrange(desc(avg_log2FC))
  p <- VlnPlot(object = s.obj, features = head(tmp.cluster.markers, 9)$gene)  
  print(p)
  cat("\n \n")
}
```

# UMAP

## All samples, grouped by clusters
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
DimPlot(object = s.obj, reduction = "INTE_UMAP", label = TRUE, label.box = TRUE, repel = TRUE, pt.size = 1)
```

## All samples, grouped by samples
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
DimPlot(object = s.obj, reduction = "INTE_UMAP", label = TRUE, label.box = TRUE, repel = TRUE, pt.size = 1, group.by = "name")
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=21, fig.height=15}
DimPlot(object = s.obj, reduction = "INTE_UMAP", label = TRUE, label.box = FALSE, repel = TRUE, pt.size = 1, split.by = "name", ncol = 2, label.size = 12) 
```

## Violin plots: Marker genes
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
list.of.marker.genes <- c("Sell", "Ccr7", "Ly6c1", "Satb1", "Bcl6", "Cxcr5", "Tbx21", "Foxp3",
                          "Rorc", "Ifit3", "Top2a", "Cenpa", "Izumo1r", "Ctla4", "Anxa2")

print("List of genes shown on Violin plots")
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
print(list.of.marker.genes)
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
VlnPlot(object = s.obj, assay = "RNA", features = list.of.marker.genes)
```

# Sub-cluster on cluster 6 by Foxp3 gene expression

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
cluster6.cells <- row.names(subset(s.obj@meta.data, s.obj@meta.data$seurat_clusters == 6))
  
  count.mtx <- GetAssayData(object = s.obj, slot = "counts", assay = "RNA")["Foxp3", ]
  count.mtx.Foxp3.pos <- count.mtx[count.mtx > 0] %>% names()
  count.mtx.Foxp3.zero <- count.mtx[count.mtx == 0] %>% names()
  
  cluster6.cells.Foxp3.pos <- intersect(count.mtx.Foxp3.pos, row.names(subset(s.obj@meta.data, s.obj@meta.data$seurat_clusters == 6)))
  cluster6.cells.Foxp3.zero <- intersect(count.mtx.Foxp3.zero, row.names(subset(s.obj@meta.data, s.obj@meta.data$seurat_clusters == 6)))
if (file.exists(file.path(path.to.03.output, sprintf("%s.filter_contaminated_cells.clusterRes_%s.subClusterFoxp3_cluster6.rds", PROJECT, cluster.resolution))) == FALSE){
  meta.data <- s.obj@meta.data %>% 
    rownames_to_column("barcode") 
  
  assign.barcode.to.cluster6.Foxp3 <- function(input.barcode){
    if (input.barcode %in% cluster6.cells.Foxp3.pos){
      return("cluster6_Foxp3_pos")
    } else if (input.barcode %in% cluster6.cells.Foxp3.zero){
      return("cluster6_Foxp3_zero")
    } else {
      cluster <- subset(meta.data, meta.data$barcode == input.barcode)$seurat_clusters
      return(sprintf("cluster%s", cluster))
    }
  }
  meta.data <- meta.data %>%
    rowwise() %>%
    mutate(cluster.foxp3 = assign.barcode.to.cluster6.Foxp3(barcode)) %>%
    column_to_rownames("barcode")
  
  meta.data <- meta.data[row.names(s.obj@meta.data), ]
  
  s.obj <- AddMetaData(object = s.obj, metadata = meta.data$cluster.foxp3, col.name = "cluster.foxp3")
  
  saveRDS(object = s.obj, file.path(path.to.03.output, sprintf("%s.filter_contaminated_cells.clusterRes_%s.subClusterFoxp3_cluster6.rds", PROJECT, cluster.resolution)))
} else {
  s.obj <- readRDS(file.path(path.to.03.output, sprintf("%s.filter_contaminated_cells.clusterRes_%s.subClusterFoxp3_cluster6.rds", PROJECT, cluster.resolution)))
}

```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}
DimPlot(object = s.obj, reduction = "INTE_UMAP", 
        cells.highlight = list(Foxp3_pos = cluster6.cells.Foxp3.pos, Foxp3_zero = cluster6.cells.Foxp3.zero), 
        cols.highlight = list("red", "blue"))
```


```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=14, fig.height=10}

DimPlot(object = s.obj, reduction = "INTE_UMAP", label = TRUE, label.box = TRUE, repel = TRUE, group.by = "cluster.foxp3") +
  ggtitle("UMAP sub-clustering by Foxp3 gene in cluster 6")
  
```

## Cluster marker genes

Applying the `wilcox` test on all genes to find the highly expressed genes in each cluster. 

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=12}
Idents(s.obj) <- "cluster.foxp3"

if (file.exists(file.path(path.to.03.output, sprintf("cluster_marker_genes_clusterRes_subcluster_Foxp3_%s.rds", cluster.resolution))) == FALSE){
  cluster.markers.Foxp3 <- FindAllMarkers(object = s.obj, assay = "RNA", test.use = "wilcox")
  cluster.markers.Foxp3 <- subset(cluster.markers.Foxp3, cluster.markers.Foxp3$p_val_adj <= 0.05 & cluster.markers.Foxp3$avg_log2FC >= 0)
  saveRDS(cluster.markers.Foxp3, file.path(path.to.03.output, sprintf("cluster_marker_genes_clusterRes_subcluster_Foxp3_%s.rds", cluster.resolution))  )
} else {
  cluster.markers.Foxp3 <- readRDS(file.path(path.to.03.output, sprintf("cluster_marker_genes_clusterRes_subcluster_Foxp3_%s.rds", cluster.resolution)))
}

```


### Heatmap: Top N up-gene in each clusters
```{r echo=FALSE, warning=FALSE, results='asis', message=FALSE, fig.width=15, fig.height=10}
topN <- 10
tmp.cluster.markers.Foxp3.up <- subset(cluster.markers.Foxp3, cluster.markers.Foxp3$avg_log2FC >= 0 & cluster.markers.Foxp3$p_val_adj <= 0.05)

selected.genes <- data.frame()
for (cluster.id in unique(tmp.cluster.markers.Foxp3.up$cluster)){
  tmp <- subset(tmp.cluster.markers.Foxp3.up, tmp.cluster.markers.Foxp3.up$cluster == cluster.id) %>% arrange(desc(avg_log2FC))
  tmp.genes <- head(tmp, topN)
  selected.genes <- rbind(selected.genes, tmp.genes)
}

p <- DoHeatmap(object = s.obj, features = selected.genes$gene) + scale_fill_viridis_c(option = "plasma")
ggsave(plot = p, filename = "Heatmap_topN_up_genes_in_all_clusters.svg", path = file.path(path.to.03.output), device = "svg", width = 28, height = 20, dpi = 300)
```


```{r echo=FALSE, warning=FALSE, results='asis', message=FALSE, fig.width=28, fig.height=20}
print(p)
```


### Tables of all cluster marker genes {.tabset}
```{r echo=FALSE, include=FALSE}
cluster.markers.Foxp3 %>% create_dt()
```

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=30, fig.height=20}

for (cluster.id in sort(unique(cluster.markers.Foxp3$cluster))){
  tmp.cluster.markers.Foxp3 <- subset(cluster.markers.Foxp3, cluster.markers.Foxp3$cluster == cluster.id) %>% arrange(desc(avg_log2FC))
  tmp.cluster.markers.Foxp3 <- tmp.cluster.markers.Foxp3 %>% mutate_if(is.numeric, round, 6) 
  
  cat(paste("\n\n#### Cluster: ", cluster.id, "##\n"))
  print( htmltools::tagList(datatable(tmp.cluster.markers.Foxp3, extensions = 'Buttons',
                                      filter = "top",
                                      options = list(dom = 'Blfrtip',
                                                     buttons = c('copy', 'csv', 'excel', 'pdf', 'print'),
                                                     lengthMenu = list(c(10,25,50,-1),
                                                                       c(10,25,50,"All")),
                                                     columnDefs = list(list(
                                                       targets = "_all",
                                                       render = JS(
                                                         "function(data, type, row, meta) {",
                                                         "return type === 'display' && data != null && data.length > 100 ?",
                                                         "'<span title=\"' + data + '\">' + data.substr(0, 100) + '...</span>' : data;",
                                                         "}")
                                                     ))
                                      ))))
  cat("\n \n")
}

```


### Top-12 cluster marker genes `FeaturePlot` {.tabset}
Here we just show the top-12 genes (ranked by `avg_log2FC`) for each cluster. 

```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=15}
for (cluster.id in sort(unique(s.obj@meta.data$cluster.foxp3))){
  DefaultAssay(s.obj) <- "RNA"
  cat(sprintf("#### Cluster %s \n", cluster.id))
  tmp.cluster.markers.Foxp3 <- subset(cluster.markers.Foxp3, cluster.markers.Foxp3$cluster == cluster.id) %>% arrange(desc(avg_log2FC))
  p <- FeaturePlot(object = s.obj, reduction = "INTE_UMAP", features = head(tmp.cluster.markers.Foxp3, 12)$gene, ncol = 3, label = TRUE, pt.size = 0.5, label.size = 5, label.color = "red")  
  print(p)
  cat("\n \n")
}
```

### Top-12 cluster marker genes `DotPlot` {.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=15}
for (cluster.id in sort(unique(s.obj@meta.data$cluster.foxp3))){
  DefaultAssay(s.obj) <- "RNA"
  cat(sprintf("#### Cluster %s \n", cluster.id))
  tmp.cluster.markers.Foxp3 <- subset(cluster.markers.Foxp3, cluster.markers.Foxp3$cluster == cluster.id) %>% arrange(desc(avg_log2FC))
  p <- DotPlot(object = s.obj, features = head(tmp.cluster.markers.Foxp3, 9)$gene)  
  print(p)
  cat("\n \n")
}
```

### Top-12 cluster marker genes `Vlnplot` {.tabset}
```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=15}
for (cluster.id in sort(unique(s.obj@meta.data$cluster.foxp3))){
  DefaultAssay(s.obj) <- "RNA"
  cat(sprintf("#### Cluster %s \n", cluster.id))
  tmp.cluster.markers.Foxp3 <- subset(cluster.markers.Foxp3, cluster.markers.Foxp3$cluster == cluster.id) %>% arrange(desc(avg_log2FC))
  p <- VlnPlot(object = s.obj, features = head(tmp.cluster.markers.Foxp3, 9)$gene)  
  print(p)
  cat("\n \n")
}
```


```{r echo=FALSE, warning=FALSE, message=FALSE, results='asis', fig.width=20, fig.height=15}

```

